---
title: "Purrr-fecting your R workflow"
subtitle: "An introduction to functional programming <br> R Day Colombia"
author: "[Beatriz Milz](https://beamilz.com)"
format: 
  rladies-revealjs:
    footer: "Slides por [Beatriz Milz](https://beamilz.com) (@BeaMilz), feito com [Quarto](https://quarto.org/docs/presentations/revealjs/index.html) e [Quarto R-Ladies Theme](https://github.com/beatrizmilz/quarto-rladies-theme#quarto-r-ladies-theme)."
    logo: "https://github.com/beatrizmilz/2023-11-rday-colombia-purrr/blob/main/_extensions/beatrizmilz/rladies/logo.png?raw=true"
incremental: false
knitr:
  opts_chunk: 
    echo: true
---

## Who am I? {.center}

```{css, echo = FALSE}
.reveal pre code {
  font-size: 1.5em !important;
}
```


-   Co-organizer [R-Ladies São Paulo](https://rladies-sp.org)

-   PhD Candidate in Environmental Sciences at the University of Sao Paulo, Brazil

-   Teacher and Consultant at [Curso-R](https://curso-r.com/)

-   Tidyverse certified instructor - Posit

-   Contact:

    -   Email: [milz.bea@gmail.com](mailto:milz.bea@gmail.com)
    -   More in my website: [beamilz.com](https://beamilz.com/about-es)
    
## My favorite tasks with R {.center}

- Create reproducible reports

- Automate tasks


# Let's go!

## Functional programming in R

-   Functional programming is a programming paradigm

-   Functions are really important in this context!

## Simple example in R

- Non-functional programming

```{r}
#| echo: true

x <- 1
y <- 2
z <- x + y
z
```
## Simple example in R

- Functional programming

```{r}
#| echo: true

add_two_numbers <- function(x, y) {
  x + y
}
add_two_numbers(1, 2)
```


## The purrr package

- A toolbox for functional programming

- Many of the functions in purrr are alternatives to `for` loops

- Code with `{purrr}` is shorter and clearer than `for` loops

- Knowing how to use `{purrr}` allows you to use the `{furrr}` package: it has the same syntax, but allows you to run code in parallel


::: footer
Source: [Hadley Wickham](https://www.youtube.com/watch?v=EGAs7zuRutY)
:::

## Package version

-  Check if the version installed is >= 1.0.0.

```{r}
#| echo: true
packageVersion("purrr")
```

- If not, install the latest version:

```{r}
#| eval: false
#| echo: true
install.packages("purrr")
```

## My favorite functions

- purrr has many functions!

- Many work similarly.

- We will focus on some today :)

    - `map()`

    - `map_vec()`

    - `list_c()`, `list_rbind()`, `list_cbind()`

    - `pluck()`

    - `set_names()`
    
- Examples on daily tasks!
    
# Let's start!

```{r}
library(tidyverse)
```
    

## `map()`

-   `map()` is the most important function in purrr

- Given an object (can be a vector or a list) we want to apply a function to each element of that object.

- It returns a list

```{r}
#| eval: false
#| echo: true
# Structure
map(vector_or_list, name_of_the_function)
```

- There are many functions that start with `map_*()` and have similar behavior

##  Simple example of `map()`

```{r}
# Vector that we want to apply the function
countries_mercosul <- c("Brazil", "Argentina",
                        "Uruguay", "Paraguay")
map(countries_mercosul, str_to_upper)
```

## What if I want a vector to be returned? {.center}


## We can use `map_vec()`!

-  `map_vec()` is a variant of `map()` that always returns a vector!

-  It is useful when you want to apply a function to each element of a vector and return a vector.

-  It is equivalent to `map() |> list_c()`

```{r}
map_vec(countries_mercosul, str_to_upper) 
```



## We also can use `list_c()`!

- `list_c()` is a function that takes a **list** and returns a **vector**

- It is useful to use after `map()`, and you want to convert the list into a vector

```{r}
map(countries_mercosul, str_to_upper) |> list_c()
```

## `pluck()`

- `pluck()` is a function that extracts a single element from a list



- We will use a simple list for the examples:

```{r}
map(countries_mercosul, str_to_upper)
```


## Example of `pluck()`

```{r}
map(countries_mercosul, str_to_upper) |> pluck(1)
map(countries_mercosul, str_to_upper) |> pluck(2)
map(countries_mercosul, str_to_upper) |> pluck(3)
map(countries_mercosul, str_to_upper) |> pluck(4)
```
## Why we don't use `[[ ]]` instead?


- With `pluck()`:

```{r}
map(countries_mercosul, str_to_upper) |> pluck(1)
```

- With  `[[ ]]`:

```{r}
#| error: true
map(countries_mercosul, str_to_upper)[[1]]
```


## Pluck is **pipeable**! `|>`

- With `pluck()`:

```{r}
map(countries_mercosul, str_to_upper) |>
  pluck(1)
```

- With  `[[ ]]`:

```{r}
#| error: true
map(countries_mercosul, str_to_upper) |> 
  .[[1]]
```


## Pluck is safer! 

- If the element does not exist, it returns `NULL` instead of an error.

- With `pluck()`:

```{r}
map(countries_mercosul, str_to_upper) |> pluck(5)
```
- With  `[[ ]]`:

```{r}
#| error: true
map(countries_mercosul, str_to_upper)[[5]]
```


## `pluck()` with nested lists

- `pluck()` can be used to extract elements from nested lists. That's where it shines!

```{r}
# TO DO!
```

## `list_rbind()`, `list_cbind()`

- `list_rbind()` and `list_cbind()` are functions that bind lists together (by row or column) and returns a tibble.

TO DO

## `set_names()`

- `set_names()` is a function that sets names to a vector or list

```{r}

```

TO DO

# Examples on my daily tasks

TO DO

# Learn more

-   [Documentation of the purrr package](https://purrr.tidyverse.org/index.html)

- - <https://purrr.tidyverse.org/reference/list_c.html>

-   [News on the purrr package: version 1.0.0](https://www.tidyverse.org/blog/2022/12/purrr-1-0-0/)

-   Video: [Hadley Wickham \| {purrr} 1.0: A complete and consistent set of tools for functions and vectors](https://www.youtube.com/watch?v=EGAs7zuRutY)

